---
title: "citibike regression"
date: '`r Sys.time()`'
output:
    html_document:
        # code_folding: hide
        number_sections: yes
        toc: yes
        toc_depth: 3
---


```{r setup, include=FALSE}
library(tidyverse)
library(modelr)
library(lubridate)
citibike_trips <- read.table("week4/trips_per_day.tsv", sep = "\t", header = TRUE)
```

```{r load-datasets}
set.seed(42)

num_days <- nrow(citibike_trips)
fraction <- 0.9
num_train <- floor(num_days * fraction)

ndx <- sample(1:num_days, num_train, replace = FALSE)

trips_train <- citibike_trips[ndx, ]
trips_test <- citibike_trips[-ndx, ]

num_folds <- 5

trips_train_folds <- trips_train %>%
    mutate(fold = (row_number() %% num_folds) + 1)

theme_set(theme_bw())
```

```{r min-base-model}
set.seed(42)
K <- 1:10

avg_err <- c()
stderr <- c()

for (k in K) {
    validate_err <- c()
    for (f in 1:num_folds) {
        model <- lm(num_trips ~ poly(tmin, k, raw = TRUE), filter(trips_train_folds, fold != f))
        val_data <- filter(trips_train_folds, fold == f)
        err <- sqrt(mean((predict(model, val_data) - val_data$num_trips)^2))
        validate_err[f] <- err
    }

    avg_err[k] <- mean(validate_err)
    stderr[k] <- sd(validate_err) / sqrt(num_folds)
}
```


``` {r ploting-data}
avg_validate_err <- avg_err
se_validate_err <- stderr

plot_data <- data.frame(K, avg_validate_err, se_validate_err)
ggplot(plot_data, aes(x = K, y = avg_validate_err)) +
    geom_pointrange(aes(
        ymin = avg_validate_err - se_validate_err,
        ymax = avg_validate_err + se_validate_err,
        color = avg_validate_err == min(avg_validate_err)
    )) +
    geom_line(color = "red") +
    scale_x_continuous(breaks = 1:12) +
    theme(legend.position = "none") +
    xlab("Polynomial Degree") +
    ylab("RMSE on validation data")
```

```{r min-base-model-extra-features}
set.seed(42)

validate_err <- c()
adjusted_R2 <- c()
for (f in 1:num_folds) {
    model <- lm(num_trips ~ poly(prcp, 2, raw = T) + poly(tmax, 8, raw = T) + sqrt(snwd) + poly(tmin, 4, raw = T), filter(trips_train_folds, fold != f))
    val_data <- filter(trips_train_folds, fold == f)
    err <- sqrt(mean((predict(model, val_data) - val_data$num_trips)^2))
    validate_err[f] <- err
    adjusted_R2[f] <- summary(model)$adj.r.squared
}

avg_adj_r <- mean(adjusted_R2)
avg_val_err <- mean(validate_err)
se_val <- sd(validate_err) / sqrt(num_folds)

avg_adj_r
avg_val_err
se_val
```

# testing snwd and snow

num_trips ~ prcp + snwd + tmax + tmin
>     avg_adj_r
[1] 0.7926796
>     avg_val_err
[1] 4767.835
>     se_val
[1] 169.8234

num_trips ~ prcp + poly(snwd, 5, raw = T) + tmax + tmin
> avg_adj_r
[1] 0.7910862
> avg_val_err
[1] 4781.634
> se_val
[1] 166.2134

num_trips ~ prcp + snwd + tmax * tmin
> avg_adj_r
[1] 0.7921217
> avg_val_err
[1] 4791.8
> se_val
[1] 174.7397


num_trips ~ prcp + snow + tmax + tmin
> avg_adj_r
[1] 0.7828797
> avg_val_err
[1] 4915.607
> se_val
[1] 192.8162

num_trips ~ prcp + snwd * poly(snow, 2, raw = T) + tmax + tmin
> avg_adj_r
[1] 0.7918337
> avg_val_err
[1] 6142.923
> se_val
[1] 922.3327

verdict get rid of snow

# setup season

``` {r turn-ymd-to-season}
trips_season <- citibike_trips |>
    mutate(season = case_when(
        month(ymd) >= 3 & month(ymd) <= 5 ~ 1,
        month(ymd) >= 6 & month(ymd) <= 8 ~ 2,
        month(ymd) >= 9 & month(ymd) <= 11 ~ 3,
        month(ymd) == 2 ~ 4,
        month(ymd) == 1 ~ 4,
        month(ymd) == 12 ~ 4,
    )) |>
    mutate(season = as.factor(season))

is.factor(trips_season$season)

set.seed(42)

num_days <- nrow(trips_season)
fraction <- 0.9
num_train <- floor(num_days * fraction)

ndx <- sample(1:num_days, num_train, replace = FALSE)

trips_train_season <- trips_season[ndx, ]
trips_test_season <- trips_season[-ndx, ]

num_folds <- 5

trips_train_season_folds <- trips_train_season %>%
    mutate(fold = (row_number() %% num_folds) + 1)
```

# validate season

```{r min-base-model-season-extra-features}
set.seed(42)

validate_err <- c()
adjusted_R2 <- c()
for (f in 1:num_folds) {
    model <- lm(num_trips ~ poly(prcp, 2, raw = T) + poly(tmax, 8, raw = T) + sqrt(snwd) + poly(tmin, 4, raw = T) + season, filter(trips_train_season_folds, fold != f))
    val_data <- filter(trips_train_season_folds, fold == f)
    err <- sqrt(mean((predict(model, val_data) - val_data$num_trips)^2))
    validate_err[f] <- err
    adjusted_R2[f] <- summary(model)$adj.r.squared
}

avg_adj_r <- mean(adjusted_R2)
avg_val_err <- mean(validate_err)
se_val <- sd(validate_err) / sqrt(num_folds)

avg_adj_r
avg_val_err
se_val
```

# setup weekends

``` {r weekday}
trips_weekends_folds <- trips_train_folds |>
    mutate(is_weekend = as.numeric(wday(ymd) == 1 | wday(ymd) == 7))

trips_weekends_test <- trips_test |>
    mutate(is_weekend = as.numeric(wday(ymd) == 1 | wday(ymd) == 7))
```

# validate weekends

``` {r min-base-model-season-extra-features}
set.seed(42)

validate_err <- c()
adjusted_R2 <- c()
for (f in 1:num_folds) {
    model <- lm(num_trips ~ poly(prcp, 2, raw = T) + poly(tmax, 8, raw = T) + sqrt(snwd) + poly(tmin, 4, raw = T) + is_weekend, filter(trips_weekends_folds, fold != f))
    val_data <- filter(trips_weekends_folds, fold == f)
    err <- sqrt(mean((predict(model, val_data) - val_data$num_trips)^2))
    validate_err[f] <- err
    adjusted_R2[f] <- summary(model)$adj.r.squared
}

avg_adj_r <- mean(adjusted_R2)
avg_val_err <- mean(validate_err)
se_val <- sd(validate_err) / sqrt(num_folds)

avg_adj_r
avg_val_err
se_val
```

# setup both weekends and season

``` {r weekday}
trips_weekend_season_folds <- trips_train_season_folds |>
    mutate(is_weekend = as.numeric(wday(ymd) == 1 | wday(ymd) == 7))

trips_weekend_season_test <- trips_test_season |>
    mutate(is_weekend = as.numeric(wday(ymd) == 1 | wday(ymd) == 7))
```

# validate weekends and season

``` {r min-base-model-season-extra-features}
set.seed(42)

validate_err <- c()
adjusted_R2 <- c()
for (f in 1:num_folds) {
    model <- lm(num_trips ~ poly(prcp, 2, raw = T) + poly(tmax, 8, raw = T) + poly(snwd, 5, raw = T) + poly(tmin, 4, raw = T) + is_weekend * season, filter(trips_weekend_season_folds, fold != f))
    val_data <- filter(trips_weekend_season_folds, fold == f)
    err <- sqrt(mean((predict(model, val_data) - val_data$num_trips)^2))
    validate_err[f] <- err
    adjusted_R2[f] <- summary(model)$adj.r.squared
}

avg_adj_r <- mean(adjusted_R2)
avg_val_err <- mean(validate_err)
se_val <- sd(validate_err) / sqrt(num_folds)

avg_adj_r
avg_val_err
se_val
```

```{r plot-final-model-date-v-num-trips}
model <- lm(num_trips ~ poly(prcp, 2, raw = T) + poly(tmax, 8, raw = T) + poly(snwd, 5, raw = T) + poly(tmin, 4, raw = T) + is_weekend * season, trips_weekend_season_folds)
trips_predictions <- trips_weekend_season_folds |>
    add_predictions(model) |>
    mutate(pred = round(pred))

trips_predictions |> ggplot(aes(x = date)) +
    geom_line(aes(y = pred, color = "pred")) +
    geom_point(aes(y = num_trips, color = "actual"))


trips_predictions |> ggplot(aes(x = pred, y = num_trips)) +
    geom_point() +
    geom_abline(color = "red", linetype = "dashed")
```

